A Transformer Model for Retrosynthesis

### 2019-04-03 papers/retrosynthesis/article.tex
one
minor changes

The advent of deep learning\cite{ChenDrugDiscovery} gave a new impulse for virtual modeling and also opened a venue for a promising set of generative methods based on Recurrent Neural Networks\cite{ErtlLSTM}, Variational Autoencoders\cite{DuvenaudVAE}, and Generative Adversarial Networks trained with reinforcement learning\cite{ORGAN,OlivecronaReinforcment}. These techniques are changing the course of QSAR studies from the observation to the invention: from a virtual screening of available compounds to direct synthesis of new candidates. Generative models can produce big sets of promising molecules and impaired with SMILES-based QSAR methods\cite{TetkoCNF} provide a strong foundation for creating highly optimized focussed libraries, but estimation of synthetic availability of these compounds is an open question though several scores based on fragmentation\cite{ErtlSaScore} and machine learning\cite{ColeySCScore} approaches have been developed. 
To synthesize a molecule, one should have a plan of a multi-step synthesis and also a set of available reactants. 
Finding an optimal combination of reactants, reactions, and conditions to obtain the compound with good yield, sufficient quality, and quantity is not a trivial task even for experts in organic chemistry. Recent advances in the computer-aided synthesis planning are reviewed in \cite{ColeyReview,EnkvinstReview,BaskinReview}.

@@

Neural sequence-to-sequence (seq2seq) approach has been recently applied for a direct reaction 
prediction task \cite{SchwallerTranslation,SchwallerTransformer} with outstanding statistical 
parameters of final models -- 90.4\% of accuracy on test set. Seq2seq modeling has been also tested on 
retrosynthesis task\cite{Pande}, but due to the complex nature of retrosynthesis itself 
and difficulty in estimating the correct predictions of reactants
\footnote{A target molecule usually can be synthesized with different reactions starting from different sets of reactants. 
The predictions of the model may be correct from organic chemist point of view but differ from 
the reactant set in ground truth. This may lead to underestimation of effectiveness of models.}, 
accuracy on the test set was moderate 37.4\% but still comparable to rule-based systems 35.4\%. 
We questioned about the possibility of improvement models for one-step retrosynthesis utilizing modern neural network architectures 
and training techniques. Applying the Transformer Model\cite{Transformer}, 
together with cyclical learning rate schedule\cite{TransformerTips}, 
resulted in a single model with accuracy 42.7\%, 
that is more than 5\% higher compare to the baseline model\cite{Pande}.

Our main contributions are:
\begin{itemize}
\item We show that for this particular task there is no need to use a vliadation dataset for early-stopping or other parameters optimization. We trained all the parameters directly from the training dataset.
\item Applying weights averaging and snapshot learning helped to train the most precise model for one-step retrosynthesis prediction. We averaged weights on 5 successive cycles of learning rate schedule.
\item Increasing the temperature while performing a beam-search procedure improves the accuracy up to 2\%. 
\end{itemize}

@@

In this study we used the same dataset of reactions as in \cite{Pande}. 
This dataset was filtered from the USPTO database\cite{Lowe} originally derived from the USA patents and 
contains 50\,000 reactions classified into 10 reaction types\cite{Schneider}. The authors\cite{Pande} further preprocessed the database by 
splitting multiple products reactions into multiple single products reactions. The resulting dataset contains 40\,029, 5\,004, and 5\,004 reactions for training, validation, and testing respectively. Information about the reaction type 
was discarded as we aimed at building a general model using SMILES of products and reactants only.

@@

\subsection{Training heuristics}
Training a Transformer model is a challenge, and several heuristics have been proposed\cite{TransformerTips}, some of them were used in this study:

\paragraph{Using as bigger batch size as possible.} However, due to our hardware limitations we could not set the batch size more then 64\footnote{Our first implementation of the model required a lot of memory to deal with masks of reactants and products. Though later we improved the code we still remained this size for consistency of the results.};
\paragraph{Iincreasing the learning rate at the beginning} of training up to warmup steps\footnote{In our implementation 1 step is equivalent to 1 batch. The number of reactions for training is 40\,029 + 5\,004, so one epoch is equal to 704 batches.}. The authors of the original Transformer paper \cite{Transformer}  used 4\,000 steps for warming. The Transformer model for reaction prediction task from \cite{SchwallerTransformer} used 8\,000 steps. We have also tried different values for warmup and eventually found that 16\,000 works well with our model.
\paragraph{Applying cyclic learning rate} schedules. This tips can generally improve any model\cite{Izmailov} through better loss landscape exploration with bigger learning rates after the optimiser felt down to some local minima. For this study we used the following scheme for learning rate calculation depending on the step:
\[
    u(step)= 
\begin{cases}
    warmup + (step \bmod cycle), & \text{if } step \ge cycle\\
    step,              & \text{otherwise}
\end{cases}
\]
where $cycle$ stands for the number of steps while the learning rate is decreasing before raising to the maximum again.
\begin{equation}
\label{eq:rate}
\lambda(step) = factor * { {min (1.0, u(step) / warmup ) } \over { max (u(step), warmup) } }
\end{equation}	
where $factor$ is just a constant. Big values of $factor$ introduce numerical instability during training, so after several trials we set $factor = 20.0$. The curve for learning rate in this study is shown in fig. \ref{fig:example}, plot (4, f). 
\paragraph{Averaging weights} during last steps (usually 10- 20) of training or at minima of learning rates in case of snapshop learning \cite{Snapshot}. Also with cyclic learning rate schedules it is possible to average weights of those models that have minimum in losses just before increasing of the rate. Such approach leads to more stable and plain region in loss landscapes \cite{Izmailov}.

\section{Results}
\label{sec:results}

\begin{wrapfigure}{r}{8cm}  
  \vspace{-0.2cm}
  \includegraphics[width = 8cm]{images/tempdep.pdf}
  \caption{Dependence of the beam search on temperature. For better exploration, higher temperatures are more useful. In this study we explored T=1.3. Bigger values significantly worse for Top-3, and approximately the same for Top-1 and Top-5. This curve was derived from the training dataset.}
  \label{fig:tempdep}
  \vspace{-0.5cm}
\end{wrapfigure}
For this study, we implemented The Transformer model in Tensorflow \cite{tensorflow} library to support its integration in our in-house programs set (\url{https://github.com/bigchem/retrosynthesis}).  All values reported are averages for three repetitive runs, see p. \pageref{sec:appendix} for details. Preliminary modeling showed that the architecture with 3 layers and 8 attention heads works well for the datasets, though we tried combinations of 2, 3, 4, 5 layers with 6, 8, 10, 12 heads. So all calculations were performed with these values fixed.  The number of learnable parameters of the model is 1\,882\,176, embedding layer common for product and reactants has size 64.
\begin{table}[b!]
  \caption{Accuracy (\%) of the models on test set when all reactants were correctly predicted.}
  \centering
  \begin{tabular}{p{2cm}p{1cm}p{1cm}p{1cm}p{1cm}p{8cm}}
    \toprule
    Model & Greedy & Top-1 & Top-3 & Top-5 & Description\\
    \midrule
    Seq2Seq &  & 37.4  & 52.4 & 57.0 & Literature result from \cite{Pande} based on Seq2Seq architecture. \\ \midrule
    T1         & 34.4 & 37.9 & 57.3 & 62.7 & Transformer Model trained with validation control set (early stopping, \textasciitilde 200 epochs). \\
    T$1_1$ & 37.3 & 39.8 & 59.1 & 63.9 & The same as T1, but without early stopping (1000 epochs). \\ 
    T2         & 39.3 & 41.8 & 61.3 & 67.2 & Transformer Model trained on both training and validation sets for 1000 epochs.\\
    \textbf{T3}         & \textbf{40.6} & \textbf{42.7} & \textbf{63.9} & \textbf{69.8} & Transformer Model trained with cyclic learning rate schedule for 1000 epochs. Averaging cycles 6,7,8,9, and 10.\\
    \bottomrule
  \end{tabular}
  \label{tbl:results}
\end{table}

@@

Then we trained our model with cyclic learning rate schedule, eq. \ref{eq:rate}, fig. \ref{fig:example} (b) for better exploration of loss landscape. During training, we also saved the character-based accuracy of the model, fig. \ref{fig:example}, (f). This snapshot training regime \cite{Snapshot} produces a set of different weights at each minimum of learning rate. Averaging them is to some extent equivalent to a consensus of models but within one model \cite{Izmailov}. We tried different averaging regimes T3 and found that averaging on 6-7-8-9-10 cycles gives better  results. 
\begin{figure}[t!]
  \centering
  \includegraphics[width = 16cm]{images/learning.pdf}
  \caption{Summary of learning curves for the Transformer model: a) original learning rate schedule with warmup; b) cyclic learning rate with warmup; c) cross-entropy loss  for training and d) validation;  e) cross-entropy loss and f) character-based accuracy for training a model wit cyclic learning schedule.}
  \label{fig:example}
\end{figure}

@@

The beam search calculates the sum of negative logarithms of probabilities of selecting a token at a particular step, and thus, this value can be a measure of internal confidence. To check this hypothesis, we selected T3-2 model and estimated its internal performance to distinguish between correct and invalid predictions.  The parameters of the classifier were: AUC = 0.77, optimal threshold = 0.00678. Then we validated the model with an additional condition: if the score is less than optimal threshold we selected the answer, otherwise we went to the next candidate in the possible reactant sets returned by the beam search. The results were even worse than without thresholds, 28.45 vs. 42.42.   A possible\begin{wrapfigure}{r}{7cm}  
  \vspace{-0.2cm}
  \includegraphics[width = 7cm]{images/auc.pdf}
  \caption{Internal classification performance.}
  \label{fig:auc}
  \vspace{-1.7cm}
\end{wrapfigure}
 explanation is that the estimation does not deal with organic chemistry. The model tries to derive some character-based scoring relying only on tokens in a string and increasing this value does not influence the quality of prognosis. The same effect we saw during training when the character accuracy is 98\% whereas chemistry-based metric is much lower.

Estimation of optimal thresholds on training sets almost always a bad idea due to the biasing of a model to its source data. The correct way is to use validation dataset instead. We built the classifier for the T1-2 with characteristics: AUC = 0.65, optimal threshold 0.00396, and applied it for testing the model. The results were again worse, 13.5\% vs. 40.85\%. There are no significant differencies of accuracies when using unnormalized or normalized on the length of the reactants string scores. Fig. \ref{fig:auc} shows ROC curves for T1-2 and T3-2 models derived at T=1.3. Evidently one cannot

>>>>

The advent of deep learning \cite{ChenDrugDiscovery} gave a new impulse for virtual modeling and also opened a venue for a promising set of generative methods based on Recurrent Neural Networks \cite{ErtlLSTM}, Variational Autoencoders \cite{DuvenaudVAE}, and Generative Adversarial Networks trained with reinforcement learning \cite{ORGAN,OlivecronaReinforcment}. These techniques are changing the course of QSAR studies from the observation to the invention: from a virtual screening of available compounds to direct synthesis of new candidates. Generative models can produce big sets of promising molecules and impaired with SMILES-based QSAR methods \cite{TetkoCNF} provide a strong foundation for creating highly optimized focussed libraries, but estimation of synthetic availability of these compounds is an open question though several scores based on fragmentation \cite{ErtlSaScore} and machine learning \cite{ColeySCScore} approaches have been developed. 
To synthesize a molecule, one should have a plan of a multi-step synthesis and also a set of available reactants. 
Finding an optimal combination of reactants, reactions, and conditions to obtain the compound with good yield, sufficient quality, and quantity is not a trivial task even for experts in organic chemistry. Recent advances in the computer-aided synthesis planning are reviewed in \cite{ColeyReview,EnkvinstReview,BaskinReview}.

@@

Neural sequence-to-sequence (seq2seq) approach has been recently applied for a direct reaction 
prediction task \cite{SchwallerTranslation,SchwallerTransformer} with outstanding statistical 
parameters of final models -- 90.4\% of accuracy on test set. Seq2seq modeling has been also tested on 
retrosynthesis task \cite{Pande}, but due to the complex nature of retrosynthesis itself 
and difficulty in estimating the correct predictions of reactants
\footnote{A target molecule usually can be synthesized with different reactions starting from different sets of reactants. 
The predictions of the model may be correct from organic chemist point of view but differ from 
the reactant set in ground truth. This may lead to underestimation of effectiveness of models.}, 
accuracy on the test set was moderate 37.4\% but still comparable to rule-based systems 35.4\%. 
We questioned about the possibility of improvement models for one-step retrosynthesis utilizing modern neural network architectures 
and training techniques. Applying the Transformer Model \cite{Transformer}, 
together with cyclical learning rate schedule \cite{TransformerTips}, 
resulted in a single model with accuracy 42.7\%, 
that is more than 5\% higher compare to the baseline model \cite{Pande}.

Our main contributions are:
\begin{itemize}
\item We show that for this particular task there is no need to use a validation dataset for early-stopping or other parameters optimization. We trained all the parameters directly from the training dataset.
\item Applying weights averaging and snapshot learning helped to train the most precise model for one-step retrosynthesis prediction. We averaged weights on 5 successive cycles of learning rate schedule.
\item Increasing the temperature while performing a beam-search procedure improves the accuracy up to 2\%. 
\end{itemize}

@@

In this study we used the same dataset of reactions as in \cite{Pande}. 
This dataset was filtered from the USPTO database \cite{Lowe} originally derived from the USA patents and 
contains 50\,000 reactions classified into 10 reaction types \cite{Schneider}. The authors \cite{Pande} further preprocessed the database by 
splitting multiple products reactions into multiple single products reactions. The resulting dataset contains 40\,029, 5\,004, and 5\,004 reactions for training, validation, and testing respectively. Information about the reaction type 
was discarded as we aimed at building a general model using SMILES of products and reactants only.

@@

\subsection{Training heuristics}
Training a Transformer model is a challenge, and several heuristics have been proposed \cite{TransformerTips}, some of them were used in this study:

\paragraph{Using as bigger batch size as possible.} However, due to our hardware limitations we could not set the batch size more then 64\footnote{Our first implementation of the model required a lot of memory to deal with masks of reactants and products. Though later we improved the code we still remained this size for consistency of the results.};
\paragraph{Iincreasing the learning rate at the beginning} of training up to warmup steps\footnote{In our implementation 1 step is equivalent to 1 batch. The number of reactions for training is 40\,029 + 5\,004, so one epoch is equal to 704 batches.}. The authors of the original Transformer paper \cite{Transformer}  used 4\,000 steps for warming. The Transformer model for reaction prediction task from \cite{SchwallerTransformer} used 8\,000 steps. We have also tried different values for warmup and eventually found that 16\,000 works well with our model.
\paragraph{Applying cyclic learning rate} schedules. This tips can generally improve any model \cite{Izmailov} through better loss landscape exploration with bigger learning rates after the optimiser felt down to some local minima. For this study we used the following scheme for learning rate calculation depending on the step:
\[
    u(step)= 
\begin{cases}
    warmup + (step \bmod cycle), & \text{if } step \ge cycle\\
    step,              & \text{otherwise}
\end{cases}
\]
where $cycle$ stands for the number of steps while the learning rate is decreasing before raising to the maximum again.
\begin{equation}
\label{eq:rate}
\lambda(step) = factor * { {min (1.0, u(step) / warmup ) } \over { max (u(step), warmup) } }
\end{equation}	
where $factor$ is just a constant. Big values of $factor$ introduce numerical instability during training, so after several trials we set $factor = 20.0$. The curve for learning rate in this study is shown in fig. \ref{fig:example}, plot (4, f). 
\paragraph{Averaging weights} during last steps (usually 10- 20) of training or at minima of learning rates in case of snapshop learning \cite{Snapshot}. Also with cyclic learning rate schedules it is possible to average weights of those models that have minimum in losses just before increasing of the rate. Such approach leads to more stable and plain region in loss landscapes \cite{Izmailov}.

\section{Results}
\label{sec:results}
\subsection {Learning details and evaluation of models for retrosynthesis prediction} 
\begin{wrapfigure}{r}{8cm}  
  \vspace{-0.5cm}
  \includegraphics[width = 8cm]{images/tempdep.pdf}
  \caption{Dependence of the beam search on temperature. For better exploration, higher temperatures are more useful. In this study we explored T=1.3. Bigger values significantly worse for Top-3, and approximately the same for Top-1 and Top-5. This curve was derived from the training dataset.}
  \label{fig:tempdep}
  \vspace{-0.5cm}
\end{wrapfigure}
For this study, we implemented The Transformer model in Tensorflow \cite{tensorflow} library to support its integration in our in-house programs set (\url{https://github.com/bigchem/retrosynthesis}).  All values reported are averages for three repetitive runs, see p. \pageref{sec:appendix} for details. Preliminary modeling showed that the architecture with 3 layers and 8 attention heads works well for the datasets, though we tried combinations of 2, 3, 4, 5 layers with 6, 8, 10, 12 heads. So all calculations were performed with these values fixed.  The number of learnable parameters of the model is 1\,882\,176, embedding layer common for product and reactants has size 64.
\begin{table}[b!]
  \caption{Accuracy (\%) of the models on test set when all reactants were correctly predicted.}
  \centering
  \begin{tabular}{p{2cm}p{1cm}p{1cm}p{1cm}p{1cm}p{8cm}}
    \toprule
    Model & Greedy & Top-1 & Top-3 & Top-5 & Description\\
    \midrule
    Seq2Seq &  & 37.4  & 52.4 & 57.0 & Literature result from \cite{Pande} based on Seq2Seq architecture. \\ \midrule
    T1         & 34.4 & 37.9 & 57.3 & 62.7 & Transformer Model trained with validation control set (early stopping, \textasciitilde 200 epochs). \\
    T$1_1$ & 37.3 & 39.8 & 59.1 & 63.9 & The same as T1, but without early stopping (1000 epochs). \\ 
    T2         & 39.3 & 41.8 & 61.3 & 67.2 & Transformer Model trained on both training and validation sets for 1000 epochs.\\
    \textbf{T3}         & \textbf{40.6} & \textbf{42.7} & \textbf{63.9} & \textbf{69.8} & Transformer Model trained with cyclic learning rate schedule for 1000 epochs. Averaging cycles 6,7,8,9, and 10.\\
    \bottomrule
  \end{tabular}
  \label{tbl:results}
\end{table}

@@

Then we trained our model with cyclic learning rate schedule, eq. \ref{eq:rate}, fig. \ref{fig:example} (b) for better exploration of loss landscape. During training, we also saved the character-based accuracy of the model, fig. \ref{fig:example}, (f). This snapshot training regime \cite{Snapshot} produces a set of different weights at each minimum of learning rate. Averaging them is to some extent equivalent to a consensus of models but within one model \cite{Izmailov}. We tried different averaging regimes for T3 and found that averaging on 6-7-8-9-10 cycles gives better  results. 
\begin{figure}[t!]
  \centering
  \includegraphics[width = 16cm]{images/learning.pdf}
  \caption{Summary of learning curves for the Transformer model: a) original learning rate schedule with warmup; b) cyclic learning rate with warmup; c) cross-entropy loss  for training and d) validation;  e) cross-entropy loss and f) character-based accuracy for training a model wit cyclic learning schedule.}
  \label{fig:example}
\end{figure}

@@

The beam search calculates the sum of negative logarithms of probabilities of selecting a token at a particular step, and thus, this value can be a measure of internal confidence. To check this hypothesis, we selected T3-2 model and estimated its internal performance to distinguish between correct and invalid predictions.  The \begin{wrapfigure}{r}{7cm}  
  \vspace{-0.2cm}
  \includegraphics[width = 7cm]{images/auc.pdf}
  \caption{Internal classification performance.}
  \label{fig:auc}
  \vspace{-0.2cm}
\end{wrapfigure}parameters of the classifier were: AUC = 0.77, optimal threshold = 0.00678. Then we validated the model with an additional condition: if the score is less than optimal threshold we selected the answer, otherwise  we went to the next candidate in the possible reactant sets returned by the beam search. The results were even worse than without thresholds, 28.45 vs. 42.42.   A possible explanation is that the estimation does not deal with organic chemistry. The model tries to derive some character-based scoring relying only on tokens in a string and increasing this value does not influence the quality of prognosis. The same effect we saw during training when the character accuracy is 98\% whereas chemistry-based metric is much lower.

Estimation of optimal thresholds on training sets almost always a bad idea due to the biasing of a model to its source data. The correct way is to use validation dataset instead. We built the classifier for the T1-2 with characteristics: AUC = 0.65, optimal threshold 0.00396, and applied it for testing the model. The results were again worse, 14.1\% vs. 40.85\%. There are no significant differencies of accuracies when using unnormalized or normalized on the length of the reactants string scores. Fig. \ref{fig:auc} shows ROC curves for T1-2 and T3-2 models derived at T=1.3. 
Evidently one cannot use this estimation to improve TOP-1 scoring.
